---
title: "RPlot_Script"
author: "Priyanka Choudhary <br /> Mike Rouw <br /> Pratheeba Nalligounder Palanisami"
date: "November 14, 2019"
output:
  html_document: 
    theme: cosmo
    toc: yes
    github_document:
      toc: yes
    word_document: default
    pdf_document:
      toc:yes
---
* * *

# Embed Code
## Set Directory

Note: Root directory path needs to be updated to your local

```{r setup, echo=FALSE, message=FALSE, warning=FALSE} 
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Priya/Desktop/RPlot") 
```

## Load data

```{r loadData}
JeanFinalDataSet <- read.csv("JeanFinalDataSet.csv")
```

# Introduction to the data plotting

(Mike)


# Duration Plot

(Intro)

The below R script is used to get the duration plot
```{r, echo=FALSE}

library(ggplot2)

ggplot(JeanFinalDataSet,aes(x=YesDurationGroup)) + 
  
  geom_bar(color="blue",fill="blue") + 
  
  labs(title="Duration of Positive Phenophase Observation", y="Number of Observations",x="Duration Groups")
```

## Analysis

(Mike)

# Species Plots

## For SpeciesDesc (sweetgum)

(Intro)

The below R script is used to get the species plot for Species Description 'Sweetgum' 

```{r, echo=FALSE}

library(ggplot2)
library(dplyr)

PlotData <- filter(JeanFinalDataSet, SpeciesDesc == "sweetgum")
ggplot(PlotData,aes(x=FirstNo, y=PhenophaseDesc, color=SiteName)) +
geom_point(aes(x=LastNo, y=PhenophaseDesc)) +
geom_point(aes(x=FirstNo, y=PhenophaseDesc)) +
geom_segment(aes(x=FirstYes, xend=LastYes, yend=PhenophaseDesc)) +
facet_grid(rows=vars(Year)) +
labs(x="Day of Year",y="Sweetgum",title="Phenophase Observations") +
theme(legend.position="bottom")

```


## Analysis

(Mike)


## For SpeciesDesc (bald cypress)

(Intro)

The below R script is used to get the species plot for Species Description 'bald cypress' 

```{r, echo=FALSE}

library(ggplot2)
library(dplyr)

PlotData <- filter(JeanFinalDataSet, SpeciesDesc == "bald cypress")
ggplot(PlotData,aes(x=FirstNo, y=PhenophaseDesc, color=SiteName)) +
geom_point(aes(x=LastNo, y=PhenophaseDesc)) +
geom_point(aes(x=FirstNo, y=PhenophaseDesc)) +
geom_segment(aes(x=FirstYes, xend=LastYes, yend=PhenophaseDesc)) +
facet_grid(rows=vars(Year)) +
labs(x="Day of Year",y="Bald cypress",title="Phenophase Observations") +
theme(legend.position="bottom")

```

## Analysis 

(Mike)

## For SpeciesDesc (trumpet creeper)

(Intro)

The below R script is used to get the species plot for Species Description 'Trumpet creeper' 

```{r, echo=FALSE}

library(ggplot2)
library(dplyr)

PlotData <- filter(JeanFinalDataSet, SpeciesDesc == "trumpet creeper")
ggplot(PlotData,aes(x=FirstNo, y=PhenophaseDesc, color=SiteName)) +
geom_point(aes(x=LastNo, y=PhenophaseDesc)) +
geom_point(aes(x=FirstNo, y=PhenophaseDesc)) +
geom_segment(aes(x=FirstYes, xend=LastYes, yend=PhenophaseDesc)) +
facet_grid(rows=vars(Year)) +
labs(x="Day of Year",y="Trumpet creeper",title="Phenophase Observations") +
theme(legend.position="bottom")

```

## Analysis 

(Mike)

# Phenophase Plots

## For PhenophaseDesc (Breaking leaf buds)

(Intro)

The below R script is used to get the species plot for Phenophase Description 'Breaking leaf buds' 

```{r, echo=FALSE}

library(ggplot2)
library(dplyr)

PlotData <- filter(JeanFinalDataSet, PhenophaseDesc == "Breaking leaf buds")
ggplot(PlotData,aes(x=FirstNo, y=SpeciesDesc, color=SiteName)) +
geom_point(aes(x=LastNo, y=SpeciesDesc)) +
geom_point(aes(x=FirstNo, y=SpeciesDesc)) +
geom_segment(aes(x=FirstYes, xend=LastYes, yend=SpeciesDesc)) +
facet_grid(rows=vars(Year)) +
labs(x="Day of Year",y="Breaking leaf buds",title="Phenophase Observations") +
theme(legend.position="bottom")

```
## Analysis 

(Mike)


## For PhenophaseDesc (Open flowers)

(Intro)

The below R script is used to get the species plot for Phenophase Description 'Open flowers' 

```{r, echo=FALSE}

library(ggplot2)
library(dplyr)

PlotData <- filter(JeanFinalDataSet, PhenophaseDesc == "Open flowers")
ggplot(PlotData,aes(x=FirstNo, y=SpeciesDesc, color=SiteName)) +
geom_point(aes(x=LastNo, y=SpeciesDesc)) +
geom_point(aes(x=FirstNo, y=SpeciesDesc)) +
geom_segment(aes(x=FirstYes, xend=LastYes, yend=SpeciesDesc)) +
facet_grid(rows=vars(Year)) +
labs(x="Day of Year",y="Open flowers",title="Phenophase Observations") +
theme(legend.position="bottom")

```
## Analysis 

(Mike)



## For PhenophaseDesc (Colored leaves)

(Intro)

The below R script is used to get the species plot for Phenophase Description 'Colored leaves' 

```{r, echo=FALSE}

library(ggplot2)
library(dplyr)

PlotData <- filter(JeanFinalDataSet, PhenophaseDesc == "Colored leaves")
ggplot(PlotData,aes(x=FirstNo, y=SpeciesDesc, color=SiteName)) +
geom_point(aes(x=LastNo, y=SpeciesDesc)) +
geom_point(aes(x=FirstNo, y=SpeciesDesc)) +
geom_segment(aes(x=FirstYes, xend=LastYes, yend=SpeciesDesc)) +
facet_grid(rows=vars(Year)) +
labs(x="Day of Year",y="Colored leaves",title="Phenophase Observations") +
theme(legend.position="bottom")

```
## Analysis 

(Mike)



## For PhenophaseDesc (Falling leaves)

(Intro)

The below R script is used to get the species plot for Phenophase Description 'Falling leaves' 

```{r, echo=FALSE}

library(ggplot2)
library(dplyr)

PlotData <- filter(JeanFinalDataSet, PhenophaseDesc == "Falling leaves")
ggplot(PlotData,aes(x=FirstNo, y=SpeciesDesc, color=SiteName)) +
geom_point(aes(x=LastNo, y=SpeciesDesc)) +
geom_point(aes(x=FirstNo, y=SpeciesDesc)) +
geom_segment(aes(x=FirstYes, xend=LastYes, yend=SpeciesDesc)) +
facet_grid(rows=vars(Year)) +
labs(x="Day of Year",y="Falling leaves",title="Phenophase Observations") +
theme(legend.position="bottom")

```
## Analysis 

(Mike)


**_Contributorship Statement_** : Priyanka Choudhary, Mike Rouw, Pratheeba Nalligounder Palanisami contributed to the R Plot and R script undertaking and the writing of this document.

**_Proofreader statement_** : I, Priyanka Choudhary, have reviewed and proofread the following work and agree that it is ready for submission.